---
title: "Final_ project_DDD"
author: "Harshan Ragavandar Ramesh (11055132)."
date: "2024-10-05"
output:
  html_document: default
  pdf_document: default
---

library("Amelia")  # For missing data visualization
library("ranger")
```{r,include=FALSE}
# Load necessary libraries
library("ggplot2")
library("dplyr")
library("caret")
library("rpart")
library("randomForest")
library(purrr)

```

#### **1. Data Loading and Initial Summary**

```{r}
# Step 1: Load the dataset
ou_data <- read.csv("D:/MERCER/Sem 3/Data Driven/Final Project/OUData.csv")
summary(ou_data)

# Step 2: Convert necessary columns to appropriate types
# Convert character variables to numeric variables
ou_data$BloodPressureUpper <- as.numeric(ou_data$BloodPressureUpper)
ou_data$BloodPressureDiff <- as.numeric(ou_data$BloodPressureDiff)
ou_data$Pulse <- as.numeric(ou_data$Pulse)
ou_data$PulseOximetry <- as.numeric(ou_data$PulseOximetry)
ou_data$Respirations <- as.numeric(ou_data$Respirations)
ou_data$Temperature <- as.numeric(ou_data$Temperature)

# Change 0 to "No" and 1 to "Yes" for Flipped variable
ou_data$Flipped <- ifelse(ou_data$Flipped == 1, "Yes", "No")

# Convert character variables to factor variables
ou_data$Gender <- as.factor(ou_data$Gender)
ou_data$PrimaryInsuranceCategory <- as.factor(ou_data$PrimaryInsuranceCategory)
ou_data$DRG01 <- as.factor(ou_data$DRG01)
ou_data$Flipped <- as.factor(ou_data$Flipped)

# Drop InitPatientClassAndFirstPostOUClass and ObservationRecordKey variables
ou_data <- subset(ou_data, select = -c(InitPatientClassAndFirstPostOUClass, ObservationRecordKey))

# Step 3: Identify and clean non-numeric values
# Check for any non-numeric values in BloodPressureUpper and OU_LOS_hrs using str() instead of unique()
str(ou_data)

# Step 4: Check for missing values
if (any(is.na(ou_data))) {
  message("Missing values found in the dataset.")
} else {
  message("No missing values found.")
}

# Step 5: Impute missing values using the mean of numerical variable columns, and using formula BloodPressureDiff = BloodPresureUpper - BloodPressureLower for BloodPressureDiff columns
ou_data$PulseOximetry[is.na(ou_data$PulseOximetry)] <- mean(ou_data$PulseOximetry, na.rm = TRUE)
ou_data$BloodPressureUpper[is.na(ou_data$BloodPressureUpper)] <- mean(ou_data$BloodPressureUpper, na.rm = TRUE)
ou_data$Pulse[is.na(ou_data$Pulse)] <- mean(ou_data$Pulse, na.rm = TRUE)
ou_data$Respirations[is.na(ou_data$Respirations)] <- mean(ou_data$Respirations, na.rm = TRUE)
ou_data$Temperature[is.na(ou_data$Temperature)] <- mean(ou_data$Temperature, na.rm = TRUE)
ou_data$BloodPressureDiff[is.na(ou_data$BloodPressureDiff)] <- mean(ou_data$BloodPressureUpper-ou_data$BloodPressureLower, na.rm = TRUE)


# Step 6: Verify the dataset is clean again
if (sum(is.na(ou_data)) == 0) {
  message("Dataset is clean with no missing values.")
} else {
  message("There are still missing values.")
}

#Summary data
summary(ou_data)

```
#### **2. Exploratory Data Analysis (EDA)**

##### **2.1 Distribution of Variables (Numerical and Categorical Variables)**
```{r}
#Step 1: Creat bar plot for categorical variables
# Bar plot for Gender
ggplot(ou_data, aes(x = Gender)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of Gender")

# Bar plot for Primary Insurance Category
ggplot(ou_data, aes(x = PrimaryInsuranceCategory)) +
  geom_bar(fill = "darkorange") +
  labs(title = "Distribution of Primary Insurance Category")

# Bar plot for Flipped (Yes/No)
ggplot(ou_data, aes(x = Flipped)) +
  geom_bar(fill = "green") +
  labs(title = "Distribution of Flipped")

# Step 2: Create histograms for numerical variables
# Create numeric_vars
numeric_vars <- ou_data[, sapply(ou_data, is.numeric)]

# Load rlang for tidy evaluation
library(rlang)

for (var in names(numeric_vars)) {
  print(
    ggplot(ou_data, aes(x = !!sym(var))) +
      geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7, na.rm = TRUE) +
      theme_minimal() +
      labs(title = paste("Histogram of", var), x = var, y = "Frequency")
  )
}

```

```{r}
# Step 3: Boxplot for numeric variables to detect outliers
for (var in names(numeric_vars)) {
  print(
    ggplot(ou_data, aes(y = !!sym(var))) + 
      geom_boxplot(fill = "coral", color = "black", na.rm = TRUE) +
      theme_minimal() +
      labs(title = paste("Boxplot of", var), y = var)
  )
}
```

##### **2.2 Remove Outliers**
```{r}
# Detect outliers using IQR for all numeric variables

# Define the outlier detection function
detect_outliers <- function(column) {
    lower_bound <- quantile(column, 0.25) - 1.5 * IQR(column, na.rm = TRUE)
  upper_bound <- quantile(column, 0.75) + 1.5 * IQR(column, na.rm = TRUE)
    column > upper_bound | column < lower_bound
}

# Apply the function to all numeric columns
outliers <- numeric_vars %>%
  map(detect_outliers) %>%  
  bind_cols() %>%  
  rowSums() %>%  
  as.logical()

# Filter the rows with outliers in any numeric column
outliers_data <- ou_data[outliers, ]

# Remove outliers
ou_data_clean<- ou_data[!outliers, ]
```

##### **2.3 Correlation and Relationships**
```{r}
# Step 1: Correlation matrix for numeric variables
numeric_corr <- cor(numeric_vars, use = "complete.obs")

# Heatmap of correlations
library(corrplot)
corrplot(numeric_corr, method = "color", type = "upper", tl.col = "black", tl.srt = 45)

# Step 2: Boxplot to compare numeric variables with Flipped decision
for (var in names(numeric_vars)) {
  print(
    ggplot(ou_data_clean, aes_string(x = "Flipped", y = var, fill = "Flipped")) +
      geom_boxplot() +
      theme_minimal() +
      labs(title = paste("Boxplot of", var, "by Flipped"), x = "Flipped", y = var))}

# Step 3: Bar plots to see how Flipped decision relates to categorical variables
ggplot(ou_data_clean, aes(x = PrimaryInsuranceCategory, fill = Flipped)) +
  geom_bar(position = "dodge") +
  labs(title = "Flipped Decision by Primary Insurance Category") +
  theme_minimal()

ggplot(ou_data_clean, aes(x = Gender, fill = Flipped)) +
  geom_bar(position = "dodge") +
  labs(title = "Flipped Decision by Gender") +
  theme_minimal()

ggplot(ou_data_clean, aes(x = DRG01, fill = Flipped)) +
  geom_bar(position = "dodge") +
  labs(title = "Flipped Decision by DRG01") +
  theme_minimal()

# Step 4: Scatter plots for relationships

# Age vs Length of Stay
ggplot(ou_data_clean, aes(x=Age, y=OU_LOS_hrs)) + 
  geom_point(alpha=0.5, color = "skyblue2") +
  geom_smooth(method = "lm", color = "coral", se = TRUE) +
  ggtitle("Age vs Length of Stay") + 
  xlab("Age") + 
  ylab("Length of Stay (Hours)")


# Age vs Oxygen Level (PulseOximetry)
ggplot(ou_data_clean, aes(x = Age, y = PulseOximetry)) +
  geom_point(color = "green3", alpha = 0.6) + 
  geom_smooth(method = "lm", color = "coral", se = TRUE) + 
  labs(title = "Age vs Oxygen Level", 
       x = "Age", y = "Oxygen Level (PulseOximetry)")

# Blood Pressure Lower vs Pulse
ggplot(ou_data_clean, aes(x=BloodPressureLower, y=Pulse)) + 
  geom_point(alpha=0.5, color = "coral") +  
  geom_smooth(method = "lm", color = "coral", se = TRUE) +
  ggtitle("Blood Pressure Lower vs Pulse") + 
  xlab("Blood Pressure Lower") + 
  ylab("Pulse")
```


#### 3. Classification (Logistic Regression, Decision Tree, and Random Forest)**

##### **3.1 Logistic Regression Model**
```{r}

# Step 1: Partition datacet into 80% of training and 20% of testing

set.seed(123)
train.index=sample(c(1:dim(ou_data_clean)[1]), dim(ou_data_clean)[1]*0.8)
train_data  = ou_data_clean[train.index, ]
test_data = ou_data_clean[-train.index, ]

# Step 2: Build the logistic regression model
log_model <- glm(Flipped ~ .-OU_LOS_hrs, 
                 data = train_data, family = binomial)
# Summary Logistic Regression                 
summary(log_model)

# Predictions
log_pred <- predict(log_model, test_data, type = "response")
log_pred_class <- ifelse(log_pred > 0.5, "Yes", "No")

# Confusion matrix
log_conf_matrix <- confusionMatrix(factor(log_pred_class), test_data$Flipped)
print(log_conf_matrix)

# Step 3: Build other the logistic regression model
log_model2 <- glm(Flipped ~ Gender + PrimaryInsuranceCategory + DRG01 , 
                 data = train_data, family = binomial)
# Summary Logistic Regression                 
summary(log_model2)           

# Predictions
log_pred2 <- predict(log_model2, test_data, type = "response")
log_pred_class2 <- ifelse(log_pred2 > 0.5, "Yes", "No")

# Confusion matrix
log_conf_matrix2 <- confusionMatrix(factor(log_pred_class2), test_data$Flipped)
print(log_conf_matrix2)
```

##### **3.2 Decision Tree Model**

```{r}
# Decision Tree
tree_model <- rpart(Flipped ~ .-OU_LOS_hrs, 
                    data = train_data, method = "class")
# Plot the tree
library(rpart.plot)
rpart.plot(tree_model, 
           main = "Decision Tree for Flipped Decision",
           type = 3,      
           extra = 100,
           fallen.leaves = TRUE, 
           digits = 2, cex=0.7)

# Predictions
tree_predictions <- predict(tree_model, test_data, type = "class")
confusionMatrix(tree_predictions, test_data$Flipped)
```


##### **3.3 Random Forest Model**

```{r}
# Random Forest Model
set.seed(123)
rf_model <- randomForest(Flipped ~ Gender+ PrimaryInsuranceCategory + DRG01 , 
                         data = train_data, ntree = 100)

# Summary
print(rf_model)

# Predictions
rf_predictions <- predict(rf_model, test_data)
confusionMatrix(rf_predictions, test_data$Flipped)
```


#### **Step 4: Model Evaluation**
```{r}
# Function to calculate performance metrics
get_metrics_with_fp_fn <- function(predictions, actual) {
  conf_matrix <- confusionMatrix(factor(predictions), factor(actual))
  
  # Extracting values from the confusion matrix
  accuracy <- conf_matrix$overall['Accuracy']
  sensitivity <- conf_matrix$byClass['Sensitivity']
  specificity <- conf_matrix$byClass['Specificity']
  
  return(data.frame(
    Accuracy = accuracy,
    Sensitivity = sensitivity,
    Specificity = specificity
  ))
}

# Get metrics for Logistic Regression Model 1
log_metrics1 <- get_metrics_with_fp_fn(log_pred_class, test_data$Flipped)

# Get metrics for Logistic Regression Model 2
log_metrics2 <- get_metrics_with_fp_fn(log_pred_class2, test_data$Flipped)

# Get metrics for Decision Tree Model
tree_metrics <- get_metrics_with_fp_fn(tree_predictions, test_data$Flipped)

# Get metrics for Random Forest Model
rf_metrics <- get_metrics_with_fp_fn(rf_predictions, test_data$Flipped)

# Combine metrics into a single data frame for comparison
model_comparison <- rbind(
  cbind(Model = "Logistic Regression 1", log_metrics1),
  cbind(Model = "Logistic Regression 2", log_metrics2),
  cbind(Model = "Decision Tree", tree_metrics),
  cbind(Model = "Random Forest", rf_metrics)
)

# Print the comparison table
print(model_comparison)


```



**Step 5: Finalize Model Selection:**

##### **5.1 Retrain the Final Model on the Entire Dataset**

```{r}
# Retrain the logistic regression model on the entire dataset
final_model <- glm(Flipped ~ .- OU_LOS_hrs, 
                         data = ou_data_clean, family = binomial)

# Display the summary of the final model to check coefficients and performance
summary(final_model)
```

##### **5.2 Analyze the Final Model’s Performance**

```{r}
# Predictions
final_pred <- predict(final_model, ou_data_clean, type = "response")
final_pred_class <- ifelse(final_pred > 0.5, "Yes", "No")

# Confusion matrix
final_conf_matrix <- confusionMatrix(factor(final_pred_class), ou_data_clean$Flipped)
print(final_conf_matrix)
```
**Step 6: Final Model for Future Use:** 

```{r}
# Save the final logistic regression model to an .RDS file for future use
saveRDS(final_model, file = "final_logistic_regression_model.rds")

# To load the model later for predictions or analysis
# finalmodel <- readRDS("final_logistic_regression_model.rds")
```


- Final Insights:

```{r}
# Extract model coefficients and interpret them
coefficients <- coef(final_model)
print(coefficients)

# You can also calculate other goodness-of-fit metrics like AIC or pseudo R-squared
AIC(final_model)  # Lower AIC is better

```

